Early warning signals: The charted and uncharted territories

### 2013-05-21 manuscript.tex
built pdf

Not all regime shifts are rapid. Some systems undergo bifurcations
between qualitatively different, but quantitatively similar regimes.
These transitions may be reversible. In a management setting, such
qualitative changes may be insignificant, so warning signals that detect
such transitions may be effective ``false positives.''

CSD precedes several types of these non-catastrophic bifurcations. In
the subcritical form of a Hopf bifurcation, a system transitions from a
stable equilibrium to a stable cycle. As a control parameter approaches
the critical threshold, the system's dominant eigenvalue approaches zero
and thus exhibits CSD (Chisholm and Filotas 2009, Kéfi et al. 2012).
However, the mean value of the equilibrium does not change dramatically,
and the transition from stable equilibrium to cycles is gradual as the
cycle sizes grow from zero at the threshold value. To appreciate how
this bifurcation is gradual rather than catastrophic, note that in the
presence of stochasticity, the system behavior observed on either side
of the threshold may be indistinguishable: on one side stochasticity
bounces the system around a stable node, while on the other it bounces
the system around a very small limit cycle in the same region of state
space. Contrast this to a critical transition in which any stochastic
fluctuation across the threshold could lead to a qualitatively different
state.

>>>>

Not all regime shifts are rapid. Some systems undergo bifurcations
between qualitatively different, but quantitatively similar regimes.
These transitions may be reversible. In a management setting, such
qualitative changes may be gradual, so warning signals that detect such
transitions may be effective ``false positives.''

CSD precedes several types of these non-catastrophic bifurcations. In
the subcritical form of a Hopf bifurcation, a system transitions from a
stable equilibrium to a stable cycle. As a control parameter approaches
the critical threshold, the system's dominant eigenvalue approaches zero
and thus exhibits CSD (Chisholm and Filotas 2009, Kéfi et al. 2012).
However, the mean value of the equilibrium does not change dramatically,
and the transition from stable equilibrium to cycles is gradual as the
cycle sizes grow from zero at the threshold value. To appreciate how
this bifurcation is gradual rather than catastrophic, note that in the
presence of stochasticity, the system behavior observed on either side
of the threshold may be indistinguishable: on one side stochasticity
bounces the system around a stable node, while on the other it bounces
the system around a very small limit cycle in the same region of state
space. Even when oscillations grow quickly, returning the environmental
conditions (bifurcation parameter) to the previous conditions restores
the stable node -- the bifurcation does not exhibit the hysteresis of
the saddle node bifurcation. Contrast this to a critical transition in
which any stochastic fluctuation across the threshold could lead to a
qualitatively different state.
